Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_

### 2020-03-17 reviewer_response.Rmd

---
title: "Response to reviewers"
subtitle: "for AMPPS manuscript: Visualisation of Brain Statistics with R-packages ggseg and ggseg3d"
author: "Athanasia Monika Mowinckel & Didac Vidal Piñeiro"
date: "2/14/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reviewer 1

### Major concerns
#### 1. The output of many code chunks are not present to illustrate the concepts described (Pages 5, 9).

The code for Figure 3 on page 5 is printed on page 5, _after_ the figure. 
This is necessary, as there is not really any proper room for the code to come before that plot specifically.
We hoped this was not too confusing, but understand that is it easy to miss.

on the plot for page 9, the missing code is printed in the title of each sub-plot. 
The reason we are omitting the code chunk of this image is that to incorporate static images of 3d plotly images, we need to use the tool `orca`. We mention this tool on page 11, but we do not want to go into orca in this tutorial. This is because it adds a level of complexity which we do not want to introduce, as we wish to focus on the bits of code necessary to make the images in an interactive R session. 
We have added an _unexectuted_ code chunk showing these three plots, but not how you can combine them into the three-part subplot that is actually in the paper, as this is just a little too much information in our opinion.

#### 2. Many users wish to bring in parcellations from individual subjects processing. For example, freesurfer files or PLY files from some other software. It is not clear if ggseg has the ability to read and then render these, but they are fundamental to this area. Similarly, In this vein, the overall functionality of a user creating their own atlas would be helpful.
We agree, this is an area we are working on imporiving. 
Currently, creating one's own custom atlas is cumbersome and not something anyone can do easily. 
While we hope to get this functionality improved in the (near) future, we currently have information on how to make your own `ggseg` atlas in the github wiki'es page. 
**ADD INFORMATION FOR ggseg3d creation!**

#### 3. In this regard, please read in a true freesurfer stats file from the subjects folder installed with freesurfer, and don’t use mock data. This will help those using Freesurfer, which is a likely user of these packages.  
We considerably altered the section introducing how to supply your own data to the function, to include this aspect of reading a freesurfer file directly.
Indeed, when we did this, we realised a bit more data manipulation was necessary to get the data into correct format than we could expect a novice user to handle.
Therefore, we added several new functions to the package, that should make this transition easier.
The section is now changed to also instruct the user on how to use those functions to read in Freesurfer data from two different formats, and these are subsequently plotted in Figure 4.

We do continue to use mock data in the remainder of the tutorial, because the mock data allows us to highlight certain function features more easily.

#### 4. The interactivity of these tools is essential - link out to interactive examples are necessary.
We have added links to online documentation of the tools, which have more features covered, and in case of `ggseg3d` also has interactive versions of plots.

#### 5. Page 9, Line 34 “It is important to note that the coordinates in the plot (X, Y, Z) are not any type of radiological coordinate system, but arbitrary Cartesian plot coordinates.” - it is unclear why this is important - please discuss the signifiance (such as using this to find ROIs on MNI space, etc).
We added a sentence to this section making sure this point was made even more clearly.

> It is important to note that the coordinates in the plot (X, Y, Z) are **not** any type of radiological coordinate system, but arbitrary Cartesian plot coordinates.
While these coordinates are important when running imaging analyses and for reporting of the placement of these regions compared to other regions, this information is not preserved in `ggseg3d`, at the moment.
As such, information regarding referecing the regions' location in standard radiological spaces (like MNI or Taliarch) should be aquired directly from imaging software or papers describin the atlases, not `ggseg3d`.

### Minor concerns
#### 1. Why must you group_by and then facet_wrap in line 43 page 7? Is the grouping necessary? Line 23, page 7 seems to indicate that it is, but it’s unclear why
This is a pretty annoying neccessity because the function `ggseg` is just a wrapper for `geom_polygon` and we are in a sense cricumventing the genious of the grammar of graphics. We added a small section in the manuscript explaining this, and also added a subplot to figure 5, now showing when data is grouped before and if not grouped. `ggseg` will merge supplied data with atlas-data, and then feed that merged data to `ggplot` and `geom_polygon`. This makes some unfortunate changes to the data, so that the entire atlas will not be plotted correctly once provided to `ggplot` and facetted. We also added some text explaining this under the **Creating subplots** heading for ggseg.

> In regards to the second rule of needing to group the data before plotting, this is due to the underlying structure of how the provided data and the atlas data are combined before plotting. 
Since the package function, so far, is a wrapper for {ggplot2} rather than a  `geom`, there is certain ggplot2-functionality that is difficult to integrate, like the facets. 
For the complete atlas-data to be merged with _every data group_ correctly, the ggseg function needs to have an indication before plotting what the groups are. 
The result of ungrouped data into `ggseg` is shown in  \@ref(fig:datasupp3)A. 
Notice how there are three panels rather than two, and that the regions _not represented in the data_ have their own panel with all the grey `NA` colour, while the panels with the data only have the regions represented in the data.

#### 2. Can a continuous mapping of triangles be handled? For example, Figure 4 has a parcellation, but can a mapping to each point on the surface be handled? If so, please give example in supplement. 
unfortunately no. We are finding it difficult to figure out how the indexing of each individual triangle actually works, and how to map that in a continuous way. We hope to be able to crack that nut in the future, but at the moment we are not able to do so. If we get there, that would be a great improvement to the package as a whole, as it would greatly minimize the data needed to plot and improve speed of plotting, which now can be quite slow for atlases with many parcellations.

#### 3. Define “glass brains” for the main audience (page 9)
We have added the following centences below the glass brain subheading

> A glass brain is a transluscent representation of the brain that can provide a frame of reference, particularly when looking at subcortical parcellations.
Since individual brain areas or subcortical structures can be very hard to place without the visual reference of the entire brain, glass brains are often used to aid the viewer understand a regions location.
Subcortical atlases include cortical surfaces and other landmark structures for visualization purposes only.

#### 4. A link to an interactive figure 9 should be provided.

#### 5. Why must title = " " in the code chunks Page (3)
This was something hacky we did to get title alignment and placement to work with cowplot. We switched to patchwork, where this is not needed. We have not made title and subtitle setting in `labs` that describe the plots very shortly.

#### 6. The code chunks in Page 3 should be run in the same order as the figure so that readers flipping back and forth do not have to search around, p1, med, combo, p2, left, p3 is the order they are passed, so they should be the order in which they are created. Also A/B/C/etc should be in the code comments, not simply the name, e.g. F: aseg default theme.
We have corrected the figure and code with your comments, and find it much improved.

#### 7. Why is there an NA in Figure 3?
All cortical atlases have `NA` for the polygon representing the medial wall, and currently subcortical atlases also have `NA` for the polygon(s) representing the cortex.  
This was a design decision we did early on, that areas that are _not_ part of the atlas itself should be represented stored as `NA` , rather than giving them their own fill colour.

### Typos
**• extra “s” in pluss in 3rd paragraph (line 55)**  
Thank you for spotting this, we amended the error.

## Reviewer 2
**I have done many R package tutorials, and I saw no issue with this one. I would rank it as intermediate in difficulty because of some setting up additional dependencies such as the remotes package. I think I would especially utilize this package with Freesurfer so the examples provided in congruence with that workflow was useful to me. Code itself was clean, commented, and easy to follow. While much of the work I do is outside of R completely, I could see this being a useful tool for someone who is basing their analyses in that framework. I hope that the contents of this tutorial are being similarly represented on a Git documentation page, as I am sure most users will default to that mode to learn about this package instead of a manuscript. I look forward to the planned improvements that the authors note elsewhere in the Planned Package Improvements portion of the article and thank them for the open-source distribution of this package.**

**I can provide to the authors minimally commented code I wrote to follow this tutorial, but would note that most of it overlaps with existing content in the manuscript.**

### 2020-03-17 msc_ggseg.Rmd
update with revisions comments

---
title: Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_
preprint: true
author: 
  - name: Athanasia M. Mowinckel
    affiliation: 1
    corresponding: true
    email: a.m.mowinckel@psykologi.uio.no
  - name: Didac Vidal-Piñeiro
    affiliation: 1
affiliation:
  - code: 1
    address: Center for Lifespan Changes in Brain and Cognition, University of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  There is an increased emphasis on visualizing neuroimaging results in more intuitive ways. Common statistical tools for dissemination, such as bar charts, lack the spatial dimension that is inherent in neuroimaging data. Here we present two packages for the statistical software R, _ggseg_ and _ggseg3d_, that integrate this spatial component. The _ggseg_ and _ggseg3d_ packages visualize pre-defined brain segmentations as both 2D polygons and 3D meshes, respectively. Both packages are integrated with other well-established R-packages, allowing great flexibility. In this tutorial, we present the main data and functions in the _ggseg_ and _ggseg3d_ packages for brain atlas visualization. The main highlighted functions are able to display brain segmentation plots in R. Further, the accompanying _ggsegExtra_-package includes a wider collection of atlases, and is intended for community-based efforts to develop more compatible atlases to _ggseg_ and _ggseg3d_. Overall, the _ggseg_-packages facilitate parcellation-based visualizations in R, improve and ease the dissemination of the results, and increase the efficiency of the workflows. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    citation_package: natbib
    fig_caption: true
    includes:
      in_header: preamble.tex
  rticles::peerj_article: preprint
---

```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, collapse = TRUE, results = 'hold', cache=TRUE, fig.pos = 'H') 
options(tinytex.verbose = TRUE, tinytex.clean = FALSE)

library(kableExtra)
library(tidyverse)
library(plotly) 
library(ggplot2)
library(knitr)
library(ggseg)
library(ggseg3d)

@@

dd <-  bind_rows(dt, dt3) %>% 
  bind_rows(dtE) %>% 
  filter(!grepl("pal", Item)) %>% 
  mutate(Type = ifelse(grepl("3d", Item), "Mesh", "Polygon"),
         Item = gsub("_3d", "", Item)) %>% 
  select(Package, Title, Type, Item) %>% 
  spread(Type, Package, fill = "None")  %>% 
  mutate(Citation = case_when(
    grepl("dkt", Item) ~ "\\citet{dkt}",
    grepl("yeo", Item) ~ "\\citet{yeo2011}",
    grepl("aseg", Item) ~ "\\citet{aseg}",
    grepl("desterieux", Item) ~ "\\citet{desterieux}",
    grepl("^chen", Item) ~ "\\citet{chen}",
    grepl("glasser", Item) ~ "\\citet{glasser}",
    grepl("icbm", Item) ~ "\\citet{icbm}",
    grepl("jhu", Item) ~ "\\citet{jhu}",
    grepl("schaefer", Item) ~ "\\citet{schaefer}",
    grepl("tracula", Item) ~ "\\citet{tracula}",
    grepl("^ho", Item) ~ "\\citet{ho}",
    TRUE ~ ""
  ))
rm(dt, dt3, dtE)
options(warnPartialMatchAttr = FALSE,
        warnPartialMatchDollar = FALSE,
        warnPartialMatchArgs = FALSE)
```

@@

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The _ggseg_ and _ggseg3d_ – pluss the complimentary _ggsegExtra_ – packages include pre-compiled data sets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in _ggseg_ is based on polygons and _ggplot2_-based grammar of graphics (@ggplot), while the 3D functionality in _ggseg3d_ is based on tri-surface mesh plots and _plotly_ (@plotly).

@@

```{r 'init', fig.width=3, message=FALSE, fig.cap='By default `ggseg()` will plot the dkt atlas in grey shaded polygons.'}
# remotes::install_github("LCBC-UiO/ggseg")
library(ggseg)
library(dplyr)
library(tidyr)

@@

```{r 'collection', fig.width=5, fig.cap='*ggseg* plots can be used with any *ggplot2* feature such as standard scales and themes. For cortical atlases, one can supply special *ggseg* options to determine, for example, hemisphere, view, or position. **A:** `dkt` atlas, stacked with dark theme ; **B:** `dkt` with medial view only; **C:** `dkt` atlas with only left medial display; **D:** `dkt` atlas, stacked, with classic theme; **E:** `dkt` atlas with left hemisphere only; **F:** complete `aseg` atlas'}
# dkt dark theme
p1 <- ggseg(position = "stacked") +
  theme_dark() +
  labs(title=" ")

# dkt classic theme
p2 <- ggseg(position = "stacked") +
  theme_classic() +
  labs(title = " ")

# dkt medial view
med <- ggseg(view = "medial") +
  labs(title = " ")

# dkt left hemisphere
left <- ggseg(hemisphere = "left") +
  labs(title = " ")

# aseg default theme
p3 <- ggseg(atlas=aseg) +
  labs(title = " ")

# dkt left medial alone
combo <- ggseg(view = "medial",
               hemisphere = "left") +
  labs(title=" ")

# Combine plots to Figure 2
cowplot::plot_grid(p1, med, combo,  p2, left, p3,
                   labels = c("A: dkt - dark", "B: dkt - medial",
                              "C: dkt - combo", "D: dkt - classic",
                              "E: dkt - left", "F: aseg"),
                   hjust = -.05)
```

@@

```{r 'fill', fig.width=5, fig.height=5, fig.cap="Supplying 'area' to the fill option in `ggseg()`, will use the column 'area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The `dkt` atlas default palette corresponds to the FreeSurferColorLut scheme."}
# Figure 3
ggseg(mapping=aes(fill = area), colour="black") +
  scale_fill_brain("dkt") +
  theme(legend.justification=c(1,0),
        legend.position="bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

@@

```{r, size='small', eval=FALSE}
# Look at the top 5 rows of the dkt dataset
head(dkt, 5)
```

In any atlas, the column 'label' is particularly useful for combining the data of interest with the _ggseg_-polygons.
The column 'label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in _ggseg_ is in a long format - that is each region has a single row - and any data of interest needs to be in this same format.
Often data sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged to work with ggseg.
See below an example of wide-to-long conversion.

```{r, size='small'}
# Create a mock freesurfer output
freesurfer_stats <- data.frame(
  id = c(10:12),
  lh_superiortemporal = c(3.32, 4.1, 3.5),
  lh_precentral = c(2.3, 2.5, 2.1),
  lh_rostralmiddlefrontal = c(3.3, 3.2, 3.1)
)
freesurfer_stats

# Wrangle wide format data into long format
freesurfer_long <- freesurfer_stats %>%
  gather(label, thickness, -id)
freesurfer_long
```

Data in long format can then be used directly with the `ggseg()`-function, as the 'label' column corresponds in name and content with the 'label' column in the atlas data of `dkt`.
The data **must** include a column that has the same column name and at least _some_ data matching the values in the corresponding column in the atlas data.
In the next example we create some data with 4 rows, with an 'area' and 'p' column, representing the results of a hypothetical analysis.
The `ggseg()`-function will recognise the matching column 'area', and merge the supplied data into the atlas using _dplyr_-joins.
We use the 'p' column as the column flooding the segment with colour.
The appearance of the plot can then be modified similarly to any other _ggplot2_ graph using functions such as scales, labs, themes, etc., as seen in Figure \@ref(fig:datasupp)

```{r 'datasupp', fig.width=5, fig.cap="Supplying data through the '.data' option in `ggseg()` enables use of columns in the supplied data to aesthetical arguments (such as 'fill'). The *ggseg* plot can be used with any other polygon compatible function from *ggplot2* or *ggplot2* extentions, for instance adding title, changing the legend name and the colour scheme with standard *ggplot2* functions."}
# Make some mock data
someData = data.frame(
  area = c("transverse temporal", "insula",
           "pre central","superior parietal"),
  p = sample(seq(0,.5,.001), 4),
  stringsAsFactors = FALSE)

# Figure 4
ggseg(.data=someData, mapping=aes(fill=p))  +
  labs(title="A nice plot title", fill="p-value") +
  scale_fill_gradient(low="firebrick",high="goldenrod")
```

@@

### Creating subplots
There is often the need to plot a statistic of interest in different groups (e.g. thickness or brain - cognition relationships in young or older adults).
This may be obtained also with `ggseg()`, using _ggplot2_'s `facet_wrap` or `facet_grid`, with three guiding rules: 
**1)** as before, data needs to be in long format with a column indexing which group the row corresponds to (group data should appear in seperate rows, not in separate columns).
**2)** The data needs to be grouped using _dplyr_'s `group_by()` function _before_ providing the data to the `ggseg()`-function.
The `ggseg()`-function will detect grouped data, and adapt it to `facet`'s requirements. 
**3)** Apply `facet_wrap` or `facet_grid` to the plot having used the above two rules. 
An example of this can be seen in Figure \@ref(fig:datasupp3), where a mock data set including summary statistics for two groups ('Young' and 'Old') is used when faceting a _ggseg_-plot.

@@

```{r 'datasupp3', fig.width=5, fig.cap='To visualize data corresponding to different groups, one needs to use `group_by` or similar **before** plotting with `ggseg()` for facetting to work.'}
# Make some mock data
someData = data.frame(
  area = rep(c("transverse temporal", "insula",
               "pre central","superior parietal"),2),
  p = sample(seq(0,.5,.001), 8),
  AgeG = c(rep("Young",4), rep("Old",4)),
  stringsAsFactors = FALSE) %>%
  group_by(AgeG)

# Figure 5
ggseg(.data=someData, colour="white", position = "stacked",
      mapping=aes(fill=p)) +
  facet_wrap(~AgeG, ncol=2) +
  theme(legend.position = "bottom")
```

```{r 'atlases', fig.width=5, fig.cap='The `aseg` atlas showing subcortical structures has some distinct differences from the `dkt`. For instance, there is no option to show only a single hemisphere. Furthermore, rather than showing lateral and medial surfaces, it shows an axial and sagittal slice.'}
# Figure 6
ggseg(atlas="aseg", mapping=aes(fill=area)) + 
  theme(legend.justification=c(1,0),
        legend.position="bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

@@

(ref:gseg3d-1-out-cap) The three surface options provided in *ggseg3d* atlases. **From left to right:** the 'white' surface is the white matter surface, 'LCBC' surface is the semi-inflated white matter surface (inflated over 10 iterations), and the 'inflated' surface is a inflated grey matter cortical surface as provided by the FreeSurfer software.

@@

The 3D-atlas data is stored in nested tibbles.
Each cortical atlas has data sets for three different surfaces (see Figure \@ref(fig:ggseg3d-1-out)) and the two hemispheres.
Only one surface is available for subcortical atlases as inflation procedures are irrelevant.
The 'ggseg_3d' column includes all necessary information for `ggseg3d()` to create a 3D mesh-plot, and should not be modified by the user.
The additional 3D-atlases in _ggsegExtra_ have the same data structure.
It is important to note that the coordinates in the plot (X, Y, Z) are **not** any type of radiological coordinate system, but arbitrary Cartesian plot coordinates.

@@

(ref:glassbrain1-cap) For subcortical structure visualization, one can add a glass brain to the plot. This will help with locating the structures relative to the cortex, and make the plot easier to interpret. The glass brain is controlled by three options: opacity, hemisphere, and colour.

>>>>

---
title: Visualisation of Brain Statistics with R-packages _ggseg_ and _ggseg3d_
preprint: true
author: 
  - name: Athanasia M. Mowinckel
    affiliation: 1
    corresponding: true
    email: a.m.mowinckel@psykologi.uio.no
  - name: Didac Vidal-Piñeiro
    affiliation: 1
affiliation:
  - code: 1
    address: Center for Lifespan Changes in Brain and Cognition, University of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  There is an increased emphasis on visualizing neuroimaging results in more intuitive ways. Common statistical tools for dissemination, such as bar charts, lack the spatial dimension that is inherent in neuroimaging data. Here we present two packages for the statistical software R, _ggseg_ and _ggseg3d_, that integrate this spatial component. The _ggseg_ and _ggseg3d_ packages visualize pre-defined brain segmentations as both 2D polygons and 3D meshes, respectively. Both packages are integrated with other well-established R-packages, allowing great flexibility. In this tutorial, we present the main data and functions in the _ggseg_ and _ggseg3d_ packages for brain atlas visualization. The main highlighted functions are able to display brain segmentation plots in R. Further, the accompanying _ggsegExtra_-package includes a wider collection of atlases, and is intended for community-based efforts to develop more compatible atlases to _ggseg_ and _ggseg3d_. Overall, the _ggseg_-packages facilitate parcellation-based visualizations in R, improve and ease the dissemination of the results, and increase the efficiency of the workflows. 
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    citation_package: natbib
    fig_caption: true
    includes:
      in_header: preamble.tex
  rticles::peerj_article: preprint
keep_tex: true
---

```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE, 
                      collapse = TRUE, results = 'hold', 
                      cache=FALSE, fig.pos = 'H') 
options(tinytex.verbose = TRUE, tinytex.clean = FALSE,
        tibble.width = 50)

library(dplyr)
library(tidyr)
library(plotly) 
library(ggplot2)
library(ggseg)
library(ggseg3d)
library(ggsegExtra)
library(kableExtra)
library(knitr)

@@

dd <-  bind_rows(dt, dt3) %>% 
  bind_rows(dtE) %>% 
  filter(!grepl("pal", Item)) %>% 
  mutate(Type = ifelse(grepl("3d", Item), "Mesh", "Polygon"),
         Item = gsub("_3d", "", Item),
         Item = gsub("_", "", Item)) %>% 
  select(Package, Title, Type, Item) %>% 
  spread(Type, Package, fill = "None")  %>% 
  mutate(Citation = case_when(
    grepl("dkt", Item) ~ "\\citet{dkt}",
    grepl("yeo", Item) ~ "\\citet{yeo2011}",
    grepl("aseg", Item) ~ "\\citet{aseg}",
    grepl("desterieux", Item) ~ "\\citet{desterieux}",
    grepl("^chen", Item) ~ "\\citet{chen}",
    grepl("glasser", Item) ~ "\\citet{glasser}",
    grepl("icbm", Item) ~ "\\citet{icbm}",
    grepl("jhu", Item) ~ "\\citet{jhu}",
    grepl("schaefer", Item) ~ "\\citet{schaefer}",
    grepl("tracula", Item) ~ "\\citet{tracula}",
    grepl("^ho", Item) ~ "\\citet{ho}",
    TRUE ~ ""
  ))
rm(dt, dt3, dtE)
options(warnPartialMatchAttr = FALSE,
        warnPartialMatchDollar = FALSE,
        warnPartialMatchArgs = FALSE)
```

@@

In this tutorial, we introduce two packages for visualizing brain atlas data in R.
The _ggseg_ and _ggseg3d_ – plus the complimentary _ggsegExtra_ – packages include pre-compiled data sets for different brain atlases that allow for 2D and 3D visualization.
The two-dimensional functionality in _ggseg_ is based on polygons and _ggplot2_-based grammar of graphics (@ggplot), while the 3D functionality in _ggseg3d_ is based on tri-surface mesh plots and _plotly_ (@plotly).

@@

```{r 'init', fig.width=3, message=FALSE, fig.cap='By default `ggseg()` will plot the dkt atlas in grey shaded polygons.'}
# remotes::install_github("LCBC-UiO/ggseg")
library(ggseg)

# package combining plots easily
library(patchwork)

@@

```{r 'collection', fig.width=5, fig.cap='*ggseg* plots can be used with any *ggplot2* feature such as standard scales and themes. For cortical atlases, one can supply special *ggseg* options to determine, for example, hemisphere, view, or position. **A:** `dkt` atlas, stacked with dark theme ; **B:** `dkt` with medial view only; **C:** `dkt` atlas with only left medial display; **D:** `dkt` atlas, stacked, with classic theme; **E:** `dkt` atlas with left hemisphere only; **F:** axial view of `aseg` atlas'}
# A: dkt dark theme
p1 <- ggseg(position = "stacked") +
  theme_dark() +
  labs(title="dkt", subtitle = "dark theme")

# B: dkt medial view
p2 <- ggseg(view = "medial") +
  labs(title = "dkt", subtitle = "medial view")

# C: dkt left medial alone
p3 <- ggseg(view = "medial",
            hemisphere = "left") +
  labs(title="dkt", subtitle = "medial left side")

# D: dkt classic theme
p4 <- ggseg(position = "stacked") +
  theme_classic() +
  labs(title = "dkt", subtitle = "classic theme")

# E: dkt left hemisphere
p5 <- ggseg(hemisphere = "left") +
  labs(title = "dkt", subtitle = "left hemisphere")

# F: aseg default theme
p6 <- ggseg(atlas = aseg, view = "axial") +
  labs(title = "aseg", subtitle = "axial view")

# Using patchwork to combine all plots into one
(p1 | p2 | p3) / (p4 | p5 | p6) +
  plot_annotation(tag_levels = 'A')
```

@@

```{r 'fill', fig.width=5, fig.height=5, fig.cap="Supplying 'area' to the fill option in `ggseg()`, will use the column 'area' from the accompanying dataset to create a discrete colour palette over the segments in the atlas. The `dkt` atlas default palette corresponds to the FreeSurferColorLut scheme."}
# Figure 3
ggseg(mapping = aes(fill = area), colour="black") +
  scale_fill_brain("dkt") +
  theme(legend.justification = c(1,0),
        legend.position = "bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

@@

```{r, size='small', eval=FALSE, codecap = "Inspect atlas data"}
# Look at the top 5 rows of the dkt dataset
head(dkt, 5)
```

In any atlas, the column 'label' is particularly useful for combining the data of interest with the _ggseg_-polygons.
The column 'label' contains the label (region) names as in the original neuroimaging software.
For example, the DKT atlas label column matches the region names from Freesurfer statistics table outputs.
Yet, the data in _ggseg_ is in a long format - that is each region has a single row - and any data of interest needs to be in this same format.
Often data sets are organized in wide format, in which subjects are represented by rows and each different data variable is represented in a separate column, and thus need to be rearranged to work with ggseg.
To ease to conversion between Freesurfer files and ggseg compatible files, we have functions to aid import to R. 
There is an extended vignette in the [online documentation]() of the various options, here we assume that that the data has been compiled into a file using Freesurfer's `aparcstats2table` or `asegstats2table` that combine parcellation metrics for all wanted subjects into a single file. 
This file also contains extra metrics like intracranial volume and total grey matter. 
These do not map onto ggseg, and as such, one should make sure to remove them from the subsequent data.

```{r, size='small'}
file_path <- "data/aparc.volume.table"

someData <- read_freesurfer_table(file_path, measure = "volume") %>% 
  # remove rows in the data that do not conain the strings "rh" or "lh"
  dplyr::filter(grepl("rh|lh", label))
someData
```

A second option is to read in atlas files directly from the entire subject directory, using regular expression to define the atlas wanted. 
If we inspect a single subjects stats directory of Freesurfer (here on default installation on macOS), we can see several parcellations to choose from.
The default aparc has a unique string ending with `aparc.stats` which we will use to read in all subject stats for that particular atlas.
This data has more columns, as it contains several metrics that can be plotted.

```{bash}
# in bash terminal
ls /Applications/freesurfer/subjects/bert/stats
```

```{r, warning=FALSE}
atlas_data <- read_atlas_files(
  subjects_dir = "/Applications/freesurfer/subjects/",
  atlas = "aparc.stats$"
)
atlas_data
```

This long format data can be used directly with the `ggseg()`-function, as the 'label' column corresponds in name and content with the 'label' column in the atlas data of `dkt`.
The data **must** include a column that has the same column name and at least _some_ data matching the values in the corresponding column in the atlas data.
In the following example, the `ggseg()`-function will recognise the matching column 'label', and merge the supplied data into the atlas using _dplyr_-joins.
The appearance of the plot can then be modified similarly to any other _ggplot2_ graph using functions such as scales, labs, themes, etc., as seen in Figure \@ref(fig:datasupp)

```{r 'datasupp', fig.width=5, fig.cap="Supplying data through the '.data' option in `ggseg()` enables use of columns in the supplied data to aesthetical arguments (such as 'fill'). The *ggseg* plot can be used with any other polygon compatible function from *ggplot2* or *ggplot2* extentions, for instance adding title, changing the legend name and the colour scheme with standard *ggplot2* functions.", warning = FALSE}

# Figure 4A
p1 <- ggseg(.data=someData, mapping=aes(fill=volume))  +
  labs(title="from read_freesurfer_table", 
       fill="volume\n(mm^3)") +
  scale_fill_gradient(low="firebrick",high="goldenrod") +
  guides(fill = guide_colourbar(barheight = 3))

# Figure 4B
p2 <- ggseg(.data=atlas_data, mapping=aes(fill=ThickAvg))  +
  labs(title="from read_atlas_files", 
       fill="Thickness\n(mean)") +
  scale_fill_gradient(low="forestgreen",high="goldenrod")+
  guides(fill = guide_colourbar(barheight = 3))

(p1 / p2) +
  plot_annotation(tag_levels = "A")
```

@@

### Creating subplots
There is often the need to plot a statistic of interest in different groups (e.g. thickness or brain - cognition relationships in young or older adults).
This may be obtained also with `ggseg()`, using _ggplot2_'s `facet_wrap` or `facet_grid`, with three guiding rules:
**1)** as before, data needs to be in long format with a column indexing which group the row corresponds to (group data should appear in seperate rows, not in separate columns).
**2)** The data needs to be grouped using _dplyr_'s `group_by()` function _before_ providing the data to the `ggseg()`-function.
The `ggseg()`-function will detect grouped data, and adapt it to `facet`'s requirements.
**3)** Apply `facet_wrap` or `facet_grid` to the plot having used the above two rules.
An example of this can be seen in Figure \@ref(fig:datasupp3) B, where a mock data set including summary statistics for two groups ('Young' and 'Old') is used when faceting a _ggseg_-plot.

In regards to the second rule of needing to group the data before plotting, this is due to the underlying structure of how the provided data and the atlas data are combined before plotting.
Since the package function, so far, is a wrapper for {ggplot2} rather than a  `geom`, there is certain ggplot2-functionality that is difficult to integrate, like the facets.
For the complete atlas-data to be merged with _every data group_ correctly, the ggseg function needs to have an indication before plotting what the groups are.
The result of ungrouped data into `ggseg` is shown in  \@ref(fig:datasupp3) A.
Notice how there are three panels rather than two, and that the regions _not represented in the data_ have their own panel with all the grey `NA` colour, while the panels with the data only have the regions represented in the data.

@@

(ref:datasupp3-cap) Examples of incorrect and correct faceting with `ggseg`. **A**: ungrouped data will facet incorrectly due to the data not being merged correctly with the atlas. **B**: grouped data will merge in the expected way and result in correctly facetted plots.

```{r 'datasupp3', fig.width=5,  warning= FALSE, fig.cap='(ref:datasupp3-cap)', out.width="100%"}
# Make some mock data
someData = tibble(
  area = rep(c("transverse temporal", "insula",
               "pre central","superior parietal"),2),
  p = sample(seq(0,.5,.001), 8),
  AgeG = c(rep("Young",4), rep("Old",4))
)

# Figure 5a
grouped_plot <- someData %>%
  ggseg(position = "stacked",
      mapping=aes(fill=p), show.legend = FALSE) +
  facet_wrap(~AgeG, ncol=3) +
  labs(title = "Ungrouped data")

# Figure 5b
ungrouped_plot <- someData %>%
  group_by(AgeG) %>%
  ggseg(position = "stacked",
      mapping=aes(fill=p), show.legend = FALSE) +
  facet_wrap(~AgeG, ncol=3) +
  labs(title = "Grouped data")

# Combining plots with patchwork, applying theme to all
(grouped_plot / ungrouped_plot) +
  plot_layout(widths = 1) + 
  plot_annotation(tag_levels = 'A') &
  theme(axis.text = element_text(size=8))
```

The `ggseg` package also has extensive [online documentation](https://lcbc-uio.github.io/ggseg/), with more functionality than we have covered here.
The online documentation is also regularly updated with explanations of new features or code improvements.

```{r 'atlases', fig.width=5, fig.cap='The `aseg` atlas showing subcortical structures has some distinct differences from the `dkt`. For instance, there is no option to show only a single hemisphere. Furthermore, rather than showing lateral and medial surfaces, it shows an axial and sagittal slice.', collapse = TRUE}
# Figure 6
ggseg(atlas="aseg", mapping=aes(fill=area)) +
  theme(legend.justification=c(1,0),
        legend.position="bottom",
        legend.text = element_text(size = 5)) +
  guides(fill = guide_legend(ncol = 3))
```

@@

(ref:gseg3d-1-out-cap) The three surface options provided in *ggseg3d* atlases. **From left to right:** the 'white' surface is the white matter surface, 'LCBC' surface is the semi-inflated white matter surface (inflated over 10 iterations), and the 'inflated' surface is a inflated grey matter cortical surface as provided by the FreeSurfer software. The title for each plot is the literal code use to make the subplot.

```{r "ggseg3d-1", message=FALSE, echo=TRUE, eval=FALSE, out.width='30%',fig.show='hold',fig.ncol = 3, fig.cap='(ref:gseg3d-1-out-cap)'}
# Figure 7 left
ggseg3d(surface = "white")  %>%
  pan_camera("right lateral") %>%
  remove_axes() %>%
  layout(title="\nggseg3d(surface='white')",
         font =  list(size = 30))

# Figure 7 center
ggseg3d()  %>%
  pan_camera("right lateral") %>%
  remove_axes() %>%
  layout(title="\nggseg3d(surface='LCBC')",
         font =  list(size = 30))

# Figure 7 right
ggseg3d(surface = "inflated") %>%
  pan_camera("right lateral") %>%
  remove_axes() %>%
  layout(title="\nggseg3d(surface='inflated')",
         font =  list(size = 30))
```

@@

The 3D-atlas data is stored in nested tibbles.
Each cortical atlas has data sets for three different surfaces (see Figure \@ref(fig:ggseg3d-1-out)) and the two hemispheres.
Only one surface is available for subcortical atlases as inflation procedures are irrelevant.
The 'ggseg_3d' column includes all necessary information for `ggseg3d()` to create a 3D mesh-plot, and should not be modified by the user.
The additional 3D-atlases in _ggsegExtra_ have the same data structure.
It is important to note that the coordinates in the plot (X, Y, Z) are **not** any type of radiological coordinate system, but arbitrary Cartesian plot coordinates.
While these coordinates are important when running imaging analyses and for reporting of the placement of these regions compared to other regions, this information is not preserved in `ggseg3d`, at the moment.
As such, information regarding referencing the regions' location in standard radiological spaces (like MNI or Taliarch) should be acquired directly from imaging software or papers describing the atlases, not `ggseg3d`.

@@

(ref:glass brain1-cap) For subcortical structure visualization, one can add a glass brain to the plot. This will help with locating the structures relative to the cortex, and make the plot easier to interpret. The glass brain is controlled by three options: opacity, hemisphere, and colour.
